The Fremont Frontier \\\large{Living at the Margins of Maize Farming}

### 2023-04-30 manuscript/manuscript.tex
🚀 initial commit

---
title: The Fremont Frontier
subtitle: Living at the Margins of Maize Farming
date: today
author:
  - name: Kenneth B. Vernon
    email: Kenneth.Vernon@colorado.edu
    affiliations: 
      - id: colorado
        address: | 
          Center for Collaborative Synthesis in Archaeology,  
          Institute of Behavioral Science,  
          University of Colorado, Boulder,  
          1440 15th Street,  
          Boulder, CO 80302
    attributes:
        corresponding: true 
  - name: Peter M. Yaworsky
    affiliations:
      - id: aarhus-culture
        address: |
          Department of Culture and Heritage Studies, 
          School of Culture and Society, 
          Aarhus University, 
          Moesgård Allé 20, building 4216, 
          8270 Højbjerg, Denmark   
      - id: aarhus-biodiversity
        address: |
          Center for Biodiversity Dynamics in a Changing World, 
          Department of Biology, 
          Aarhus University, 
          Ny Munkegade 114-116, 
          8000 Aarhus C, Denmark
  - name: Weston McCool
    affiliations:
      - id: utah-anthropology
        address: | 
          University of Utah Department of Anthropology, 
          260 S. Central Campus Drive, RM 4625, 
          Salt Lake City, UT 84112
  - name: Jerry D. Spangler
    affiliations:
      - ref: utah-anthropology
  - name: Simon Brewer
    affiliations:
      - id: utah-geography
        address: |
          University of Utah Department of Geography, 
          260 S. Central Campus Drive, RM 4625, 
          Salt Lake City, UT 84112
      - id: gcsc
        address: |
          University of Utah Global Change and Sustainability Center, 
          115 South 1460 East, RM 234 FASB, 
          Salt Lake City, UT 84112
  - name: Brian F. Codding
    affiliations:
      - ref: utah-anthropology
      - ref: gcsc
abstract: |
  The Fremont provide an important case study to examine the resilience of ancient farmers to climatic downturns, for they lived at the far northern margin of intensive maize agriculture in the American West, where the constraints on maize production are made abundantly clear. Using a tree-ring and simulation-based reconstruction of average annual precipitation and maize growing degree days, along with cost-distance to perennial streams, we model spatial variability in Fremont site density in the eastern Great Basin. The results of our analysis have implications for defining the ecological envelope in which farming is a viable strategy across this arid region and can be used to predict where and why maize farming strategies might evolve and eventually collapse as climate changes over time.
keywords:
  - Ideal Free Distribution
  - Great Basin
  - Paleoclimate Reconstruction
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Quaternary International
      formatting: preprint
      model: 3p
execute: 
  echo: false
  error: false
  warning: false
---

# Introduction

Archaeological populations of subsistence maize farmers known collectively as "the Fremont" lived at the far northern periphery of maize farming in western North America, in an area encompassing much of the modern state of Utah north of the Virgin and Colorado Rivers, from roughly 2000 to 700 years BP [@madsen1989; @madsen1998]. That the Fremont managed to thrive for as long as they did in this part of the world poses an interesting question, for conditions there are extremely harsh and inhospitable for maize farming. Archaeologists interested in that question have typically focused on strategies directly related to farming and subsistence, like irrigation methods, trade, and seasonal high elevation hunting [e.g., @barlow2008; @boomgarden2019; @janetski2002; @hart2021; @madsen1998; @metcalfe1985; @patterson2010; @spangler1993; @morgan2012]. Here, however, we come at the problem from a slightly different angle. We seek to understand not so much *what* they did to make farming effective, but *where* they chose to do it, and why they chose those places over others [e.g., @bocinsky2014; @thomson2020]. Those are not entirely separate questions, of course, as they trade-off each other, but we think a geographic analysis can actually help illuminate some of the costs and benefits of the other strategies adopted by the Fremont. A geographic or spatial model will also help us to investigate the ways that climate change might have structured those costs and benefits, and by extension the settlement decisions of Fremont farmers, which we consider a long-term record of human adaptation to climate change.

As a test case, we focus on those Fremont living in the eastern Great Basin of western Utah, which we refer to as the western Fremont (see @fig-overview). The extreme desert conditions that the western Fremont adapted to provide a useful backdrop for exploring Fremont settlement, for variation in that environment exhibits abrupt and quite dramatic changes over short distances of space and time, thus heightening the differences between areas the western Fremont did and did not occupy. We explore those differences from a socio-ecological perspective, focusing on dynamic interactions between competing individuals on the one hand and individuals and their environment on the other [@bird2006; @kennett2006a]. This dynamic is captured nicely by the Ideal Free Distribution model from Behavioral Ecology, as it represents individual decisions about where to live as a choice between habitats whose benefits to the individual may be constrained by what other individuals are doing. The technical term for these benefits is 'suitability', which the IFD defines as some function of both the habitat's intrinsic or pristine environmental quality and its current population size [@fretwell1969; @codding2015; @winterhalder2010]. The fundamental assumption underlying the IFD is that individuals will behave optimally, that they will choose to settle the habitat with the highest suitability first, meaning the one that offers them the greatest ratio of benefits to costs relative to the available alternatives.

Assuming that western Fremont individuals were adept at responding to variation in their local ecology, in particular, that they optimized their choice of habitats, we can use an empirical or inductive species distribution model of Fremont sites to reconstruct the ecological niche for maize farming [@vernon2021; @yaworsky2020; @elith2009]. This is sometimes referred to as ecological niche modeling. The basic idea here is that we can look at the distribution of Fremont sites as a reliable indicator of the sorts of available environmental conditions that would best promote maize farming in an arid landscape. In effect, we are helping ourselves to a little reverse engineering of the archaeological record [@dennett1995]. We are assuming that whatever strategy - whatever habitat - the Fremont chose was the best alternative available to them, not in any absolute sense, of course, just the best given the constraints and trade-offs Fremont farmers faced in those environments [@jochim2022]. In other words, we are treating their settlement decisions themselves as maize farming adaptations.

![
On the left is an overview map showing the project area with modern satellite imagery. Dark green colors represent high elevation, high precipitation areas. The large beige area in the northwest is the Bonneville Basin. What remains of the Great Salt Lake can be seen just west of Salt Lake City. For visualization purposes, the map on the right shows the relative density of archaeological sites (see the section on site data below for details on how we calculate relative density). Please note that although the map appears to show many empty watersheds, most do in fact contain archaeological sites, just at extremely low densities.
](../figures/overview.png){#fig-overview fig-align="center" width="6in"}

# Background

## The Fremont

Archaeologists have nominated a number of artifact types and artifact properties as candidate traits to distinguish the Fremont from their neighbors in space and time, including metates with secondary grinding surfaces, large semi-subterranean pithouses, various clay figurines, trapezoidal rock art, one-rod-and-bundle basketry, painted white and corrugated gray ceramics, and elongated corner-notched arrow points [@madsen1998]. None of these, however, nor any combination of them, has yet proven up to the task of bringing these diverse people under a single, all-encompassing definition [@madsen1998]. Nevertheless, the Fremont do stand out, especially at their population apex roughly 1000 years BP.

Around this time, the Fremont likely reached their greatest geographic extent, inhabiting an area that encompassed most of the modern state of Utah [@janetski2011], with varying levels of support for occupations in Idaho [@dean1992prehistoric], Nevada [@hockett1998], Colorado [@baker1999], and Wyoming [@hakiel1987archery; @smith1992fremont]. These modern political boundaries roughly overlap with the eastern Great Basin and Colorado Plateau physiographic provinces.

![
Linear responses of average water-year precipitation (mm), average growing-season maize gdd (C), and cost-distance to streams (hours) to changes in elevation (m). See methods for how these values were calculated and results for more details.
](../figures/elevation-everything.png){#fig-elevation fig-align="center" width="4in"}

A basin and range topography dominates the landscape of the Great Basin, with large, dry valleys and endorheic basins punctuated by north-south trending mountain ranges. Changes in elevation in this setting are quite dramatic, on the order of several thousand meters over a short east-west transect. As elevation drives virtually every climatological process in the region, the local ecology is characterized by equally dramatic extremes [@billings1951; @grayson2011]. As shown in @fig-elevation, net water-year precipitation in the eastern Great Basin tends to increase with elevation, while maize growing degree days (GDD) over the growing-season tends to decrease. An important trade-off thus exists between temperature and precipitation, with warmer and drier conditions at lower elevations and cooler and wetter conditions at higher elevations.

These are the environments into which the Fremont first emerged in western Utah. Models for describing their origins, while diverse, tend to fall into the familiar categories of migration and diffusion. Where models of the former migration variety insist that Ancestral Puebloan farmers migrated northward from the US Southwest, bringing with them drought-tolerant maize varieties and maize farming [@madsen1975; @kidder1924], models of the latter diffusion variety suggest that the Fremont first emerged within these areas as local Archaic populations transitioned away from mobile foraging, adopting semi-sedentary agricultural practices from the southwest in a piecemeal fashion over several hundred years [@jennings1978; @winter1974]. While archaeologists have marshaled an impressive array of arguments and counter-arguments for these two classes of models, a general consensus appears to be coalescing around the idea that the truth lies somewhere in between, that the interaction between a nascent Archaic population existing at extremely low population levels and a burgeoning community of migrant farmers together produced the complex we now call the Fremont [@patterson2015; @simms1986; @spangler1993; @spangler2000; @spangler2013].

Models that try to explain *why* Fremont maize farmers emerged in the far northern periphery of the US Southwest tend to agree - more or less - that at some point farming became a more economical alternative to foraging in the region. Where they differ, it is in whether things got better for farming starting roughly 2000 years BP [@benson2011; @coltrain2002; @matson1988] or worse for foraging [@barlow2002; @broughton2010; @cannon2000; @cannon2001]. In support of the *farming-more-better* model, archaeologists draw on evidence for a suite of favorable climatic conditions that came together around 1600 years BP, including warmer temperatures and increased precipitation (both winter and summer), as well as the expansion of grasslands [@grayson2006; @hemphill1995; @rhode2000]. The greatest expansion of the Fremont, in fact, occurred around 1000 years BP, around the time the Medieval Climate Anomaly (MCA) introduced warmer summer temperatures that would have increased the intensity of summer monsoons in the southwest, pushing them further north into the Great Basin and onto the Colorado Plateau [@grayson2006], a pattern reflected in numerous tree ring records [e.g., @graybill1990; @leavitt1994; @stine1990; @salzer2014], though some complications do exist for this suggested pattern [@hart2021].

In response to these sorts of arguments, those who favor the *foraging-more-worse* model [@hart2021; @barlow2002] argue that conditions promoting maize farming would also tend to make foraging a more productive strategy. Plus, farming is an extremely costly endeavor, especially when compared to foraging, including lots of upfront investment of time and energy, so improving conditions favorable to maize agriculture are not by themselves sufficient to explain its origins. To fill that lacuna, archaeologists point to accumulating evidence suggesting that the Fremont transition to agriculture accompanied periods of sustained population growth [@barlow2008; @codding2021; @spangler2013] that likely resulted in the depressed availability of wild resources [@barlow2002; @janetski1997; @simms1986]. So, demographic pressure reduced the efficiency of foraging strategies, thus making maize agriculture the optimal strategy. And that, in turn, led the Fremont to adopt farming as their primary mode of subsistence.

Once they entered the farming niche, the Fremont quickly developed several adaptations for coping with the marginal and stochastic conditions in the eastern Great Basin and the Colorado Plateau, the most important almost certainly being irrigation (for more on this, see the discussion). On the Colorado Plateau, the eastern Fremont settled deep within the steep canyons and washes that flanked the tributaries of major rivers like the Green and Yampa [@yaworsky2021; see also, Yaworsky this issue], in places like Nine Mile Canyon [@spangler1993] and Range Creek Canyon [@hart2021]. In the eastern Great Basin, the western Fremont chose to settle on open, alluvial fans and stream terraces along the lower slopes of most ranges [@janetski2000; @madsen1998; @simms2008], at places like Nephi Mounds [@sharrock1967], Pharo Village [@marwitt1968], Median Village [@marwitt1970], Evans Mound [@berry1972; @dodd1982], and Snake Rock Village [@aikens1967].

Given the extreme aridity of the west, they were still susceptible to drought, of course, even in those more suitable areas, with a single short growing season making agricultural returns highly variable. To cope with these uncertainties, Fremont farmers may have adopted additional risk-mitigation strategies like spreading farm plots among different micro-environments, a technique otherwise known as plot diversification [@spangler1993; @spangler2013; @patterson2015]. The Fremont also relied heavily on crop storage both to provide a surplus during the winter season and to make-up for poor yields [@spangler2019].

## The Ideal Free Distribution

In population ecology, settlement decisions are represented as a choice between habitats, where each habitat $i$ has some net biological value or benefit, known as its suitability ($S_i$). Typically, though not always, this value exhibits negative density dependence, declining as a function of population size ($N_i$), perhaps owing to increased competition or resource depression. When $N_i$ is very small, the suitability of a habitat approaches its pristine or intrinsic quality ($Q_i$), or its gross value independent of all settlement costs [@greene2001]. Together, these parameters suggest a very simple model of habitat suitability:

$$
S_i = Q_i - f(N_i)
$$

This is known as the Ideal Free Distribution (IFD) model [@fretwell1969]. The IFD model assumes that individuals will choose habitats that maximize $S_i$, settling the most suitable habitat first then shifting to the next most suitable habitat when those become equal, and thereafter infilling each at an equal rate. The aggregate effect is a population distributed at equilibrium, meaning suitability is equal for all habitats. When conditions are ideal (individuals have perfect knowledge) and free (individuals face no additional costs to settlement), this distribution will also satisfy the "input matching" rule, with populations occurring at larger numbers in habitats with greater intrinsic value [@parker1978].

Two circumstances provide important exceptions to this simple model. The first concerns economies of scale, or situations in which $S_i$ exhibits positive density dependence. In ecology, these are known as Allee effects, and they are assumed to occur at low density [@allee1949; @fretwell1969]. The second concerns additional settlement costs, or costs associated with moving from one habitat to another. This is a violation of the free element of the IFD. When it occurs, the result is an ideal despotic distribution (IDD) [@fretwell1969]. This has the consequence of reducing the total population in the more suitable habitat at equilibrium.

Many social and ecological variables may be relevant to defining a general measure of $S_i$, but here we are focusing on ecological determinants of intrinsic habitat quality, $Q_i$, that are specific to maize farmers. Our reasoning here is fairly straightforward. If we assume that the distribution of Fremont populations across habitats is at equilibrium, then we can combine this concept of maize-specific suitability with the input matching rule to reconstruct the ranking of habitats on the basis of ecologically-relevant parameters. That is, it follows from their being at equilibrium that any observed differences in population size must be the result of differences in $Q_i$, that $Q_i \approx f(N_i)$ at equilibrium. So, if we have some proxy for $N_i$, like the distribution of Fremont sites across habitats, we can model that value as a function of ecological covariates $X_i$, then use the resulting model to reconstruct $Q_i \approx f(X_i)$, giving us the maize niche for Fremont farmers, both in geographic and ecological space.

# Materials and Methods

## The Project Area

Because we are focusing on the western Fremont, we constrain the current project area to the eastern Great Basin in western Utah, as shown in @fig-overview, where the Great Basin is defined in the hydrologic sense using watershed boundaries [@grayson2011]. On the far eastern edge of this region, a series of north-south ranges, including the Wasatch, Pahvant, and Tushar Mountains, together stand as the primary physiographic boundaries separating the Great Basin from the Colorado Plateau. On its far western edge, a series of north-south ranges straddle the Nevada state line or lie just inside it, including the Snake Range where Great Basin National Park is located. Its northern periphery is the Great Salt Lake and the West Desert, its southern boundary the Pine Valley Mountains that separate the Great Basin from the Virgin River watershed, just north of St. George, UT. This is a region approximately 107,000 km$^2$ in area, encompassing virtually all of the Bonneville Basin in western Utah.

These boundaries also serve as geographic borders between the western Fremont and other contemporaneous populations. Around the Great Salt Lake and the Utah-Nevada border, Fremont farming gives way to Archaic populations engaged in intensive foraging modes of subsistence. The Wasatch Range and Tushar Mountains separate the western from the eastern Fremont on the Colorado Plateau, and the Pine Valley Mountains serve as a border between the Fremont and the Virgin Ancestral Puebloan to the south [@simms2008].

## The Unit of Analysis

The unit of analysis for this research is the HUC10 watershed (n = 183), as defined by the Watershed Boundary Dataset (WBD) developed by the US Geological Survey and the Natural Resources Conservation Service within the US Department of Agriculture [@usgs2013]. These provide a spatially-explicit proxy for specifically human habitats in the proposed study area. There are two reasons for this. First, the topography of the region is such that the cost of travel between watersheds is often quite significant. As a consequence, individuals are expected to spend more time traveling within watersheds than between them, all else being equal. Second, watersheds are by definition water sinks, funneling all available run-off into their respective stream networks. Given the extreme aridity of the proposed project area, this also makes them critical resource sinks, especially for maize farmers, as watersheds determine how much water might accumulate at a location, as well as its potential for irrigation.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  number,
  preprint,
  3p]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\journal{Quaternary International}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The Fremont Frontier},
  pdfauthor={Kenneth B. Vernon; Peter M. Yaworsky; Weston McCool; Jerry D. Spangler; Simon Brewer; Brian F. Codding},
  pdfkeywords={Ideal Free Distribution, Great Basin, Paleoclimate
Reconstruction},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\setlength{\parindent}{6pt}
\begin{document}

\begin{frontmatter}
\title{The Fremont Frontier \\\large{Living at the Margins of Maize
Farming} }
\author[1]{Kenneth B. Vernon%
\corref{cor1}%
}
 \ead{Kenneth.Vernon@colorado.edu} 
\author[2,3]{Peter M. Yaworsky%
%
}

\author[4]{Weston McCool%
%
}

\author[4]{Jerry D. Spangler%
%
}

\author[5,6]{Simon Brewer%
%
}

\author[4,6]{Brian F. Codding%
%
}

\affiliation[1]{organization={},addressline={Center for Collaborative
Synthesis in Archaeology, Institute of Behavioral Science, University of
Colorado, Boulder, 1440 15th Street, Boulder, CO 80302},postcodesep={}}
\affiliation[2]{organization={},addressline={Department of Culture and
Heritage Studies, School of Culture and Society, Aarhus University,
Moesgård Allé 20, building 4216, 8270 Højbjerg, Denmark},postcodesep={}}
\affiliation[3]{organization={},addressline={Center for Biodiversity
Dynamics in a Changing World, Department of Biology, Aarhus University,
Ny Munkegade 114-116, 8000 Aarhus C, Denmark},postcodesep={}}
\affiliation[4]{organization={},addressline={University of Utah
Department of Anthropology, 260 S. Central Campus Drive, RM 4625, Salt
Lake City, UT 84112},postcodesep={}}
\affiliation[5]{organization={},addressline={University of Utah
Department of Geography, 260 S. Central Campus Drive, RM 4625, Salt Lake
City, UT 84112},postcodesep={}}
\affiliation[6]{organization={},addressline={University of Utah Global
Change and Sustainability Center, 115 South 1460 East, RM 234 FASB, Salt
Lake City, UT 84112},postcodesep={}}

\cortext[cor1]{Corresponding author}

\begin{abstract}
The Fremont provide an important case study to examine the resilience of
ancient farmers to climatic downturns, for they lived at the far
northern margin of intensive maize agriculture in the American West,
where the constraints on maize production are made abundantly clear.
Using a tree-ring and simulation-based reconstruction of average annual
precipitation and maize growing degree days, along with cost-distance to
perennial streams, we model spatial variability in Fremont site density
in the eastern Great Basin. The results of our analysis have
implications for defining the ecological envelope in which farming is a
viable strategy across this arid region and can be used to predict where
and why maize farming strategies might evolve and eventually collapse as
climate changes over time.
\end{abstract}

\begin{keyword}
    Ideal Free Distribution \sep Great Basin \sep 
    Paleoclimate Reconstruction
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, breakable, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Archaeological populations of subsistence maize farmers known
collectively as ``the Fremont'' lived at the far northern periphery of
maize farming in western North America, in an area encompassing much of
the modern state of Utah north of the Virgin and Colorado Rivers, from
roughly 2000 to 700 years BP \citep{madsen1989, madsen1998}. That the
Fremont managed to thrive for as long as they did in this part of the
world poses an interesting question, for conditions there are extremely
harsh and inhospitable for maize farming. Archaeologists interested in
that question have typically focused on strategies directly related to
farming and subsistence, like irrigation methods, trade, and seasonal
high elevation hunting
\citep[e.g.,][]{barlow2008, boomgarden2019, janetski2002, hart2021, madsen1998, metcalfe1985, patterson2010, spangler1993, morgan2012}.
Here, however, we come at the problem from a slightly different angle.
We seek to understand not so much \emph{what} they did to make farming
effective, but \emph{where} they chose to do it, and why they chose
those places over others \citep[e.g.,][]{bocinsky2014, thomson2020}.
Those are not entirely separate questions, of course, as they trade-off
each other, but we think a geographic analysis can actually help
illuminate some of the costs and benefits of the other strategies
adopted by the Fremont. A geographic or spatial model will also help us
to investigate the ways that climate change might have structured those
costs and benefits, and by extension the settlement decisions of Fremont
farmers, which we consider a long-term record of human adaptation to
climate change.

As a test case, we focus on those Fremont living in the eastern Great
Basin of western Utah, which we refer to as the western Fremont (see
Figure~\ref{fig-overview}). The extreme desert conditions that the
western Fremont adapted to provide a useful backdrop for exploring
Fremont settlement, for variation in that environment exhibits abrupt
and quite dramatic changes over short distances of space and time, thus
heightening the differences between areas the western Fremont did and
did not occupy. We explore those differences from a socio-ecological
perspective, focusing on dynamic interactions between competing
individuals on the one hand and individuals and their environment on the
other \citep{bird2006, kennett2006a}. This dynamic is captured nicely by
the Ideal Free Distribution model from Behavioral Ecology, as it
represents individual decisions about where to live as a choice between
habitats whose benefits to the individual may be constrained by what
other individuals are doing. The technical term for these benefits is
`suitability', which the IFD defines as some function of both the
habitat's intrinsic or pristine environmental quality and its current
population size \citep{fretwell1969, codding2015, winterhalder2010}. The
fundamental assumption underlying the IFD is that individuals will
behave optimally, that they will choose to settle the habitat with the
highest suitability first, meaning the one that offers them the greatest
ratio of benefits to costs relative to the available alternatives.

Assuming that western Fremont individuals were adept at responding to
variation in their local ecology, in particular, that they optimized
their choice of habitats, we can use an empirical or inductive species
distribution model of Fremont sites to reconstruct the ecological niche
for maize farming \citep{vernon2021, yaworsky2020, elith2009}. This is
sometimes referred to as ecological niche modeling. The basic idea here
is that we can look at the distribution of Fremont sites as a reliable
indicator of the sorts of available environmental conditions that would
best promote maize farming in an arid landscape. In effect, we are
helping ourselves to a little reverse engineering of the archaeological
record \citep{dennett1995}. We are assuming that whatever strategy -
whatever habitat - the Fremont chose was the best alternative available
to them, not in any absolute sense, of course, just the best given the
constraints and trade-offs Fremont farmers faced in those environments
\citep{jochim2022}. In other words, we are treating their settlement
decisions themselves as maize farming adaptations.

\begin{figure}

{\centering \includegraphics[width=6in,height=\textheight]{../figures/overview.png}

}

\caption{\label{fig-overview}On the left is an overview map showing the
project area with modern satellite imagery. Dark green colors represent
high elevation, high precipitation areas. The large beige area in the
northwest is the Bonneville Basin. What remains of the Great Salt Lake
can be seen just west of Salt Lake City. For visualization purposes, the
map on the right shows the relative density of archaeological sites (see
the section on site data below for details on how we calculate relative
density). Please note that although the map appears to show many empty
watersheds, most do in fact contain archaeological sites, just at
extremely low densities.}

\end{figure}

\hypertarget{background}{%
\section{Background}\label{background}}

\hypertarget{the-fremont}{%
\subsection{The Fremont}\label{the-fremont}}

Archaeologists have nominated a number of artifact types and artifact
properties as candidate traits to distinguish the Fremont from their
neighbors in space and time, including metates with secondary grinding
surfaces, large semi-subterranean pithouses, various clay figurines,
trapezoidal rock art, one-rod-and-bundle basketry, painted white and
corrugated gray ceramics, and elongated corner-notched arrow points
\citep{madsen1998}. None of these, however, nor any combination of them,
has yet proven up to the task of bringing these diverse people under a
single, all-encompassing definition \citep{madsen1998}. Nevertheless,
the Fremont do stand out, especially at their population apex roughly
1000 years BP.

Around this time, the Fremont likely reached their greatest geographic
extent, inhabiting an area that encompassed most of the modern state of
Utah \citep{janetski2011}, with varying levels of support for
occupations in Idaho \citep{dean1992prehistoric}, Nevada
\citep{hockett1998}, Colorado \citep{baker1999}, and Wyoming
\citep{hakiel1987archery, smith1992fremont}. These modern political
boundaries roughly overlap with the eastern Great Basin and Colorado
Plateau physiographic provinces.

\begin{figure}

{\centering \includegraphics[width=4in,height=\textheight]{../figures/elevation-everything.png}

}

\caption{\label{fig-elevation}Linear responses of average water-year
precipitation (mm), average growing-season maize gdd (C), and
cost-distance to streams (hours) to changes in elevation (m). See
methods for how these values were calculated and results for more
details.}

\end{figure}

A basin and range topography dominates the landscape of the Great Basin,
with large, dry valleys and endorheic basins punctuated by north-south
trending mountain ranges. Changes in elevation in this setting are quite
dramatic, on the order of several thousand meters over a short east-west
transect. As elevation drives virtually every climatological process in
the region, the local ecology is characterized by equally dramatic
extremes \citep{billings1951, grayson2011}. As shown in
Figure~\ref{fig-elevation}, net water-year precipitation in the eastern
Great Basin tends to increase with elevation, while maize growing degree
days (GDD) over the growing-season tends to decrease. An important
trade-off thus exists between temperature and precipitation, with warmer
and drier conditions at lower elevations and cooler and wetter
conditions at higher elevations.

These are the environments into which the Fremont first emerged in
western Utah. Models for describing their origins, while diverse, tend
to fall into the familiar categories of migration and diffusion. Where
models of the former migration variety insist that Ancestral Puebloan
farmers migrated northward from the US Southwest, bringing with them
drought-tolerant maize varieties and maize farming
\citep{madsen1975, kidder1924}, models of the latter diffusion variety
suggest that the Fremont first emerged within these areas as local
Archaic populations transitioned away from mobile foraging, adopting
semi-sedentary agricultural practices from the southwest in a piecemeal
fashion over several hundred years \citep{jennings1978, winter1974}.
While archaeologists have marshaled an impressive array of arguments and
counter-arguments for these two classes of models, a general consensus
appears to be coalescing around the idea that the truth lies somewhere
in between, that the interaction between a nascent Archaic population
existing at extremely low population levels and a burgeoning community
of migrant farmers together produced the complex we now call the Fremont
\citep{patterson2015, simms1986, spangler1993, spangler2000, spangler2013}.

Models that try to explain \emph{why} Fremont maize farmers emerged in
the far northern periphery of the US Southwest tend to agree - more or
less - that at some point farming became a more economical alternative
to foraging in the region. Where they differ, it is in whether things
got better for farming starting roughly 2000 years BP
\citep{benson2011, coltrain2002, matson1988} or worse for foraging
\citep{barlow2002, broughton2010, cannon2000, cannon2001}. In support of
the \emph{farming-more-better} model, archaeologists draw on evidence
for a suite of favorable climatic conditions that came together around
1600 years BP, including warmer temperatures and increased precipitation
(both winter and summer), as well as the expansion of grasslands
\citep{grayson2006, hemphill1995, rhode2000}. The greatest expansion of
the Fremont, in fact, occurred around 1000 years BP, around the time the
Medieval Climate Anomaly (MCA) introduced warmer summer temperatures
that would have increased the intensity of summer monsoons in the
southwest, pushing them further north into the Great Basin and onto the
Colorado Plateau \citep{grayson2006}, a pattern reflected in numerous
tree ring records
\citep[e.g.,][]{graybill1990, leavitt1994, stine1990, salzer2014},
though some complications do exist for this suggested pattern
\citep{hart2021}.

In response to these sorts of arguments, those who favor the
\emph{foraging-more-worse} model \citep{hart2021, barlow2002} argue that
conditions promoting maize farming would also tend to make foraging a
more productive strategy. Plus, farming is an extremely costly endeavor,
especially when compared to foraging, including lots of upfront
investment of time and energy, so improving conditions favorable to
maize agriculture are not by themselves sufficient to explain its
origins. To fill that lacuna, archaeologists point to accumulating
evidence suggesting that the Fremont transition to agriculture
accompanied periods of sustained population growth
\citep{barlow2008, codding2021, spangler2013} that likely resulted in
the depressed availability of wild resources
\citep{barlow2002, janetski1997, simms1986}. So, demographic pressure
reduced the efficiency of foraging strategies, thus making maize
agriculture the optimal strategy. And that, in turn, led the Fremont to
adopt farming as their primary mode of subsistence.

Once they entered the farming niche, the Fremont quickly developed
several adaptations for coping with the marginal and stochastic
conditions in the eastern Great Basin and the Colorado Plateau, the most
important almost certainly being irrigation (for more on this, see the
discussion). On the Colorado Plateau, the eastern Fremont settled deep
within the steep canyons and washes that flanked the tributaries of
major rivers like the Green and Yampa {[}\citep{yaworsky2021}; see also,
Yaworsky this issue{]}, in places like Nine Mile Canyon
\citep{spangler1993} and Range Creek Canyon \citep{hart2021}. In the
eastern Great Basin, the western Fremont chose to settle on open,
alluvial fans and stream terraces along the lower slopes of most ranges
\citep{janetski2000, madsen1998, simms2008}, at places like Nephi Mounds
\citep{sharrock1967}, Pharo Village \citep{marwitt1968}, Median Village
\citep{marwitt1970}, Evans Mound \citep{berry1972, dodd1982}, and Snake
Rock Village \citep{aikens1967}.
